A large leading tail of atmospheric escape in the HAT-P-67 b system

### 2023-04-19 paper/paper1/ms.tex
minor

\subsection{Gaia DR3}\label{gaiaDR3}
\citet{2017AJ....153..211Z} previously derived stellar radius estimates of 2.1-2.7 $R_\odot$ through Spectral Energy Distribution (SED) isochrone fitting as part of a joint orbit fit.  We systematically increase those stellar radius estimates by 8.7\% to match the greater Gaia \emph{DR3} distance (\S \ref{gaiadr3}).  For a fixed $R_p/R_\star$, the larger $R_\star$ implies a proportionally larger planet radius. This update systematically decreases the estimate for the already-low density of HAT-P-67 by 28\%, to a mere 0.035 g$\;$cm$^{-3}$, albeit with significant uncertainties from the weak mass constraint.

@@

\subsubsection{Stellar rotation rate from periodogram analysis}

@@

\subsection{Revised system evolutionary state}\label{secMISTtracks}
As discussed in Section \ref{gaiaDR3}, the slightly increased distance and radius imply a slightly more luminous host star than previously estimated.  Figure \ref{fig:evolTracks} shows evolutionary tracks using the solar-metallicity MESA \citep{2011ApJS..192....3P,2013ApJS..208....4P,2015ApJS..220...15P} Isochrones and Stellar Tracks \citep[MIST;]{2016ApJS..222....8D,2016ApJ...823..102C} for \hatp.  The tracks show that a 1.8 Gyr, 1.64 $M_\odot$ \hatp~ could be at the end of its main sequence lifetime (teal track), yielding a gradual rise in incident radiation to \hatpb.  Alternatively, a 1.54 $M_\odot$ evolutionary track would send \hatp~ along the subgiant branch at an age over 2.2 Gyr, with a rapid 50\% increase in luminosity, potentially leading to ``re-inflation'' of the planet \citep{2021ApJ...909L..16T}.  Either evolutionary track appears consistent with the observed SED, stellar surface gravity, and available constraints from spectral fitting.
\begin{figure*}
    \centering
    \includegraphics[width=0.7\linewidth]{figures/MIST_tracks_HATP.png}
    \caption{MIST evolutionary model tracks for \hatp.  The $T_\mathrm{eff}$ and luminosity of \hatp~ (gray circle) are consistent with either the late main sequence or recently evolved subgiant.}
    \label{fig:evolTracks}
\end{figure*}

@@

Excess is confidently detected at \censor{X} phases at \censor{Z}$\sigma$.  The full-width at half maximum (FWHM) of the feature is about 1.8-2.6~\AA, equivalent to a line-of-sight velocity distribution in the range of 50-75~km/s.  Individual fits to the He feature show a typical line center uncertainty $\pm0.1$~\AA~ or better.  The out-of-transit phases exhibit a line center position of 10833.4~\AA, consistent with the stellar \ion{He}{1} rest frame zero velocity.  The excess absorption resides slightly blueward of the stellar zero restframe velocity reference.  As mentioned previously, the telluric masking near 10836 \AA~ censors our view of the existence (or not) of any redshifted lobe in this region.  A more careful treatment of the telluric absorption could rescue some of the data in this region.  The excess detections exhibit a bulk blueshift of up to 30~km/s.

@@

\section{Discussion}
\subsection{Cause for stellar modulation}
As discussed in Section \censor{x}, the \emph{TESS} lightcurve exhibits up to $0.36\%$ peak-to-valley modulation amplitude, with a characteristic timescale of $p=4.7-5.9$ days. Some sectors exhibit lower amplitudes.  The conventional interpretation would consign these cyclical modulations to the familiar stellar activity: surface features---either starspots, faculae, or plage---entering and exiting the projected stellar disk on the stellar rotation period comparable to the 4.8-day orbital period of planet b.  The cause for such orbital and rotational synchronization would then ostensibly be star-planet tidal interactions.  We examine the prospect of tides in the next subsection.

>>>>

\subsection{Gaia DR3}\label{analysisgaiaDR3}
\citet{2017AJ....153..211Z} previously derived stellar radius estimates of 2.1-2.7 $R_\odot$ through Spectral Energy Distribution (SED) isochrone fitting as part of a joint orbit fit.  We systematically increase those stellar radius estimates by 8.7\% to match the greater Gaia \emph{DR3} distance (\S \ref{gaiadr3}).  For a fixed $R_p/R_\star$, the larger $R_\star$ implies a proportionally larger planet radius. This update systematically decreases the estimate for the already-low density of HAT-P-67 by 28\%, to a mere 0.035 g$\;$cm$^{-3}$, albeit with significant uncertainties from the weak mass constraint.

@@

\subsubsection{Stellar rotation rate from periodogram analysis}\label{TESSmodulation}

@@

\subsection{Revised system evolutionary state}\label{secMISTtracks}
As discussed in Section \ref{analysisgaiaDR3}, the slightly increased distance and radius imply a slightly more luminous host star than previously estimated.  Figure \ref{fig:evolTracks} shows evolutionary tracks using the solar-metallicity MESA \citep{2011ApJS..192....3P,2013ApJS..208....4P,2015ApJS..220...15P} Isochrones and Stellar Tracks \citep[MIST;][]{2016ApJS..222....8D,2016ApJ...823..102C} for HAT-P-67.  The tracks show that a 1.8 Gyr, 1.64 $M_\odot$ \hatp~ could be at the end of its main sequence lifetime (teal track), yielding a gradual rise in incident radiation to \hatpb.  Alternatively, a 1.54 $M_\odot$ evolutionary track would send \hatp~ along the subgiant branch at an age over 2.2 Gyr, with a rapid 50\% increase in luminosity, potentially leading to ``re-inflation'' of the planet \citep{2021ApJ...909L..16T}.  Either evolutionary track appears consistent with the observed SED, stellar surface gravity, and available constraints from spectral fitting.

\begin{figure*}
    \centering
    \includegraphics[width=0.7\linewidth]{figures/MIST_tracks_HATP.png}
    \caption{MIST evolutionary model tracks for \hatp.  The $T_\mathrm{eff}$ and luminosity of \hatp~ (gray circle) are consistent with either the late main sequence or recently evolved subgiant.}
    \label{fig:evolTracks}
\end{figure*}

@@

Excess is confidently detected at from phases $-$0.3 to $+$0.03.  The full-width at half maximum (FWHM) of the feature is about 1.8-2.6~\AA, equivalent to a line-of-sight velocity distribution in the range of 50-75~km/s.  Individual fits to the He feature show a typical line center uncertainty $\pm0.1$~\AA~ or better.  The out-of-transit phases exhibit a line center position of 10833.4~\AA, consistent with the stellar \ion{He}{1} rest frame zero velocity.  The excess absorption resides slightly blueward of the stellar zero restframe velocity reference.  As mentioned previously, the telluric masking near 10836 \AA~ censors our view of the existence (or not) of any redshifted lobe in this region.  A more careful treatment of the telluric absorption could rescue some of the data in this region.  The excess detections exhibit a bulk blueshift of up to 30~km/s.

@@

\section{Discussion}
\subsection{Orbital constraints}
We premised our joint \emph{TESS} lightcurve and HPF RV time series orbit fit on a circular orbit.

\subsection{Cause for stellar modulation}
As discussed in Section \ref{TESSmodulation}, the \emph{TESS} lightcurve exhibits up to $0.36\%$ peak-to-valley modulation amplitude, with a characteristic timescale of $P=4.7-5.9$ days. Some sectors exhibit lower amplitudes.  The conventional interpretation would consign these cyclical modulations to the familiar stellar activity: surface features---either starspots, faculae, or plage---entering and exiting the projected stellar disk on the stellar rotation period comparable to the 4.8-day orbital period of planet b.  The cause for such orbital and rotational synchronization would then ostensibly be star-planet tidal interactions.  We examine the prospect of tides in the next subsection.
